---
title: "tidyr"
author: "Jamie Afflerbach"
date: "10/23/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)

## read in gapminder data in wide format
gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')
```

## tidyr::gather()

```{r}
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values,
         -continent, -country)
```

## tidyr::separate()

```{r}
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values,
         -continent, -country) %>% 
  separate(col = obstype_year,
           into = c("obs_type", "year"),
           sep = "_",
           convert = TRUE)
```

## plot long data

```{r}
canada_df <- gap_long %>% 
  filter(country == "Canada",
         obs_type == "lifeExp")

ggplot(data = canada_df, aes(x = year, y = obs_values)) +
  geom_line()
```

Plot the life expectancy of all countries in the Americas

```{r}
life_df <- gap_long %>% 
  filter(obs_type == "lifeExp",
         continent == "Americas")

ggplot(data = life_df, aes(x = year, y = obs_values, color = country)) +
  geom_line()
```

## exercise

Using gap_long, calculate and plot the the mean life expectancy for each continent over time from 1982 to 2007. Give your plot a title and assign x and y labels. Hint: use the dplyr::group_by() and dplyr::summarize() functions.

```{r}
continents <- gap_long %>%
  filter(obs_type == "lifeExp",
         year > 1980) %>% 
  group_by(continent, year) %>%
  summarize(mean_le = mean(obs_values)) %>% 
  ungroup()

#plot using ggplot
ggplot(data = continents, aes(x = year, y = mean_le, color = continent)) +
  geom_line() +
  labs(x = "Year",
       y = "Age (years)",
       title = "Mean life expectancy") +
  theme_bw()
```

## spread()

```{r}
gap_normal <- gap_long %>% 
  spread(key = obs_type,
         value = obs_values)
```

## Exercise

7.6 Exercise

Convert “gap_long” all the way back to gap_wide. Hint: you’ll need to create appropriate labels for all our new variables (time*metric combinations) with the opposite of separate: tidyr::unite().

Knit the R Markdown file and sync to Github (pull, stage, commit, push)

```{r}
gap_wide_new <- gap_long %>% 
  unite(col = var_names, obs_type, year, sep = "_") %>% 
  spread(key = var_names,
         value = obs_values)
```















